---
title: "2022.03.Workbook"
output: html_document
date:  "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#  {.tabset}

## Purpose

This is the motivation (from Jacobson et al 2016):
"We quantified water quality effects of land use by calculating the relative amount of disturbed land within a watershed. A simple, yet direct watershed disturbance variable (percentage of urban, agriculture, and mining land uses in a catchment) was developed by Cross and Jacobson (2013) using National Land Cover Database 2001 land use GIS data. The percentage land use disturbance variable was significant in models predicting total phosphorus concentrations in Minnesota lakes (Cross and Jacobson 2013). Catchments with undisturbed land uses lie primarily in the Northern Forests ecoregion (CEC 1997; Level 1) and generally provide good water quality to lakes and streams in that region (Figure 5). Catchments within the agricultural Great Plains ecoregion have the highest disturbed land uses and appreciably poorer water quality (Heiskary et al. 1987; Ramstack et al. 2004). Catchments in the transition from forest to prairie in the Eastern Temperate Forests ecoregion have a wide range of disturbance values. Cross and Jacobson (2013) noted that phosphorus concentrations generally become elevated when watershed land use disturbance reached 25% and greatly increased when land use disturbances exceeded 60%. These disturbance values set the foundation for the identification of appropriate management strategies for water quality in lakes under this framework. Lakes with relatively undisturbed watersheds need protection, whereas lakes with heavily disturbed watersheds need restoration."


coding reference: https://rdrr.io/cran/mgcViz/man/plot.gamViz.html

GAM reference: http://r.qcbs.ca/workshop08/book-en/gam-with-interaction-terms.html


## Correlations

Parameters:

1. Median, natural log, epilimnetic, Total Phosphorus

2. W:L (although as noted, this is % open water which isn't truely W:L)

3. Depth (when available)

4. NLCD watershed Disturbance (2016 NLCD: VALUE_21+VALUE_22+VALUE_23+VALUE_24+VALUE_82+VALUE_81 --> #note that the 2013 article did not use 81 (pasture hay) but we include it)

5. Soil with high rate of water transmission (group A): https://www.arcgis.com/home/item.html?id=56d328b000274fcfb17864a7bb0ccaff

"Soils with parent material formed by glacial outwash are usually well drained. We used maps developed by Cum- mins and Grigal (1981) and digitized by MNDNR (2011) to identify areas with glacial outwash, which we summarized as a percent of the total watershed for each lake. We summarized land use categories assigned by the National Land Cover Database (Homer et al. 2004) as a percent of the total watershed area for each lake." - Jacobson 2013

We used group A as an alternative to this parameter used by Jacobson.

6. NLCD disturbance in the 100m buffer (2019 NLCD)

7. Septic density within the 200m buffer

```{r, echo=FALSE, results='asis', message=FALSE, warning=FALSE}
  # setwd("C:/Users/amonion/New York State Office of Information Technology Services/BWAM - Lakes Database/Current")
  # source("new_database/Reading.LMAS.Data.R")
   setwd("C:/Users/amonion/OneDrive - New York State Office of Information Technology Services/Rscripts/TP_Landuse")

  rm(list=setdiff(ls(), c("newdata","status")))

```



```{r, echo=FALSE, results='asis', message=FALSE, warning=FALSE}

#restrict to lakes 
restricted<-newdata %>%
  filter(LAKE_WATERBODY_TYPE=="LAKE") %>% 
  select(LAKE_HISTORY_ID) %>% distinct()
df<-newdata%>% 
  filter(LAKE_HISTORY_ID %in% c(unique(restricted$LAKE_HISTORY_ID))) %>% distinct() %>%  
  filter(CHARACTERISTIC_NAME %in% c("PHOSPHORUS, TOTAL"),
         substr(SAMPLE_DATE,1,4)>'2010',
         INFORMATION_TYPE=="OW",
         !is.na(LAKE_HISTORY_ID)) %>% 
  filter(RSLT_VALIDATOR_QUALIFIER!="R"|is.na(RSLT_VALIDATOR_QUALIFIER)) %>% 
  mutate(RSLT_RESULT_VALUE=ifelse(!is.na(RSLT_VALIDATOR_QUALIFIER)&RSLT_VALIDATOR_QUALIFIER=="U",RSLT_QUANTITATION_LIMIT/2,RSLT_RESULT_VALUE))%>%  
  filter(!is.na(RSLT_RESULT_VALUE)) %>% 
  select(LAKE_HISTORY_ID,SAMPLE_DATE,RSLT_RESULT_VALUE) %>% distinct()


lakedepth <- newdata%>%
  filter(LAKE_HISTORY_ID %in% c(unique(df$LAKE_HISTORY_ID))) %>% distinct() %>%
  select(LAKE_HISTORY_ID,CSLAPFD_SITE_SOUND_DEPTH,RSLT_PROFILE_DEPTH,LCIFD_MAX_SOUND_DEPTH,LCIFD_SITE_SOUND_DEPTH) %>% distinct() %>%
  mutate(CSLAPFD_SITE_SOUND_DEPTH=gsub("m","",CSLAPFD_SITE_SOUND_DEPTH),
         CSLAPFD_SITE_SOUND_DEPTH=gsub(" ","",CSLAPFD_SITE_SOUND_DEPTH),
         RSLT_PROFILE_DEPTH=gsub("m","",RSLT_PROFILE_DEPTH),
         RSLT_PROFILE_DEPTH=gsub(" ","",RSLT_PROFILE_DEPTH),
         LCIFD_MAX_SOUND_DEPTH=gsub("m","",LCIFD_MAX_SOUND_DEPTH),
         LCIFD_MAX_SOUND_DEPTH=gsub(" ","",LCIFD_MAX_SOUND_DEPTH),
         LCIFD_SITE_SOUND_DEPTH=gsub("m","",LCIFD_SITE_SOUND_DEPTH),
         LCIFD_SITE_SOUND_DEPTH=gsub(" ","",LCIFD_SITE_SOUND_DEPTH)) %>% 
  mutate(CSLAPFD_SITE_SOUND_DEPTH=as.numeric(CSLAPFD_SITE_SOUND_DEPTH),
         RSLT_PROFILE_DEPTH=as.numeric(RSLT_PROFILE_DEPTH),
         LCIFD_MAX_SOUND_DEPTH=as.numeric(LCIFD_MAX_SOUND_DEPTH),
         LCIFD_SITE_SOUND_DEPTH=as.numeric(LCIFD_SITE_SOUND_DEPTH)) %>% 
  group_by(LAKE_HISTORY_ID) %>%
  summarize(CSLAPFD_SITE_SOUND_DEPTH=max(CSLAPFD_SITE_SOUND_DEPTH,na.rm = TRUE),
            LCIFD_MAX_SOUND_DEPTH=max(LCIFD_MAX_SOUND_DEPTH,na.rm = TRUE),
            LCIFD_SITE_SOUND_DEPTH=max(LCIFD_SITE_SOUND_DEPTH,na.rm = TRUE),
            RSLT_PROFILE_DEPTH=max(RSLT_PROFILE_DEPTH,na.rm = TRUE)) %>%
   ungroup() %>%
  gather(type,depth,-LAKE_HISTORY_ID) %>% 
  mutate(depth=as.numeric(depth)) %>% 
  group_by(LAKE_HISTORY_ID) %>% 
  summarise(depth=max(depth)) %>% 
  ungroup() %>% 
  mutate(depth=ifelse(depth==-Inf,NA,depth))

df<-merge(df,lakedepth,by=c('LAKE_HISTORY_ID'),all.x=TRUE)

df<-df %>% 
  #convert to log transformed ppb
  mutate(phosphorus=1000*RSLT_RESULT_VALUE,
         depth=as.numeric(depth)) %>% 
  #unrealistic values
  filter(phosphorus==is.na(phosphorus)|phosphorus<1000) %>% 
  group_by(LAKE_HISTORY_ID) %>% 
  summarize(depth=max(depth),
         phosphorus=median(phosphorus)) %>% 
  ungroup()%>% 
  select(LAKE_HISTORY_ID,depth,phosphorus) %>% distinct() %>% 
  filter(!is.na(phosphorus)) %>% 
  #NATURAL LOG
  mutate(logTP=log(phosphorus))

#add watershed disturbance
#excluding only the lake water
landcover <- read.csv("NLCD_2019_Lake_Watershed.csv")
landcover<-landcover %>% 
  mutate_at(c('Open.Water','Developed..Open.Space', 'Developed..Low.Intensity','Developed..Medium.Intensity.','Developed..High.Intensity','Barren.Land','Deciduous.Forest','Evergreen.Forest','Mixed.Forest','Shrub.Scrub',
  'Herbaceous','Hay.Pasture','Cultivated.Crops','Woody.Wetlands','Emergent.Herbaceous.Wetlands'),as.numeric) %>% 
  mutate(#note that the 2013 article did not use 81 (pasture hay) but we include it
         totalall=Developed..Open.Space+Developed..Low.Intensity+Developed..Medium.Intensity.+Developed..High.Intensity+
           Barren.Land+Deciduous.Forest+Evergreen.Forest+Mixed.Forest+Shrub.Scrub+Herbaceous+
           Hay.Pasture+Cultivated.Crops+Woody.Wetlands+Emergent.Herbaceous.Wetlands,
         DISTURBANCE=Developed..Open.Space+Developed..Low.Intensity+Developed..Medium.Intensity.+Developed..High.Intensity+
           Hay.Pasture+Cultivated.Crops) %>% 
  rename(LAKE_HISTORY_ID = LAKE_ID) %>% 
  filter(LAKE_HISTORY_ID!=" ") %>% 
  distinct(LAKE_HISTORY_ID,.keep_all = TRUE) %>% 
  select(LAKE_HISTORY_ID,DISTURBANCE,totalall) %>% distinct() 
#add lake/watershed area
WLarea<- read.csv("NLCD_watershed_lake_ratio.csv") 
WLarea<-WLarea %>% 
  rename(LAKE_HISTORY_ID=LAKE_ID) %>% 
  select(LAKE_HISTORY_ID,WaterbodyAreaSQM,WatershedAreaSQM) %>% 
  mutate(WLarea=WatershedAreaSQM/WaterbodyAreaSQM) %>% distinct(LAKE_HISTORY_ID,.keep_all = TRUE)
landcover<-merge(landcover,WLarea,by=c('LAKE_HISTORY_ID'),all.x=TRUE)
landcover<-landcover %>% 
  #convert to a percentage
  filter(!is.na(LAKE_HISTORY_ID),!is.na(DISTURBANCE),!is.na(WaterbodyAreaSQM),!is.na(WLarea)) %>% 
  select(LAKE_HISTORY_ID,DISTURBANCE,WaterbodyAreaSQM,WLarea,totalall) %>% distinct()

#adding 100m buffer disturbance
buffer<-read.csv("NLCD_2019_Watershed_100m_buffer.csv")
buffer<-buffer %>% 
    mutate_at(c('Open.Water','Developed.Open.Space','Developed.Low.Intensity','Developed.Medium.Intensity','Developed.High.Intensity','Barren.Land','Deciduous.Forest','Evergreen.Forest','Mixed.Forest','Shrub.Scrub',
  'Herbaceous','Hay.Pasture','Cultivated.Crops','Woody.Wetlands','Emergent.Herbaceous.Wetlands'),as.numeric) %>% 
  rename(LAKE_HISTORY_ID=LAKE_ID) %>% 
  mutate(total100=(Developed.Open.Space+Developed.Low.Intensity+
           Developed.Medium.Intensity+Developed.High.Intensity+
           Barren.Land+Deciduous.Forest+Evergreen.Forest+Mixed.Forest+Shrub.Scrub+Herbaceous+
           Cultivated.Crops+Hay.Pasture+Woody.Wetlands+Emergent.Herbaceous.Wetlands),
         DISTURBANCE100m=(Developed.Open.Space+Developed.Low.Intensity+
           Developed.Medium.Intensity+Developed.High.Intensity+
           Cultivated.Crops+Hay.Pasture),
         DISTURBANCE100m=if_else(is.na(DISTURBANCE100m),0,DISTURBANCE100m)) %>% 
  select(LAKE_HISTORY_ID,DISTURBANCE100m,total100) %>% distinct()
landcover<-merge(landcover,buffer,by=c('LAKE_HISTORY_ID'),all.x=TRUE)

#adding septics
septic<-read.csv("septics.csv")
septic<-septic %>% rename(LAKE_HISTORY_ID=LAKE_ID) %>% 
  select(LAKE_HISTORY_ID,Count.of.Points) %>% distinct() %>%
  rename(septic=Count.of.Points)
landcover<-merge(landcover,septic,by=c('LAKE_HISTORY_ID'),all.x=TRUE)

landcover<-landcover %>% 
  mutate(#I added this to see if I could make 100m significant by subtracting it from the total. It didn't change conclusions below
         # DISTURBANCE=DISTURBANCE-DISTURBANCE100m,
         # totalall=totalall-total100,
         DISTURBANCE = DISTURBANCE/totalall,
         DISTURBANCE100m=DISTURBANCE100m/total100
         ) %>%
  filter(!is.na(LAKE_HISTORY_ID),!is.na(DISTURBANCE),!is.na(WaterbodyAreaSQM),!is.na(WLarea),!is.na(DISTURBANCE100m),!is.na(septic)) %>% 
  select(LAKE_HISTORY_ID,DISTURBANCE,WaterbodyAreaSQM,WLarea,DISTURBANCE100m,septic) %>% distinct()

#adding soil type
soil<-read.csv("SSURGO_2019_Lake_Watershed_soils.csv")
soil<-soil %>%  
  mutate(well_drained_soils=(GROUP_A+GROUP_B)/(GROUP_A+GROUP_B+GROUP_C+GROUP_D+GROUP_A_D+GROUP_B_D+GROUP_C_D)) %>% 
  select(LAKE_ID,well_drained_soils) %>% distinct() %>% 
  rename(LAKE_HISTORY_ID=LAKE_ID)
landcover<-merge(landcover,soil,by=c('LAKE_HISTORY_ID'),all.x=TRUE)
landcover<-landcover %>% 
  filter(!is.na(LAKE_HISTORY_ID),!is.na(DISTURBANCE),!is.na(WaterbodyAreaSQM),!is.na(WLarea),!is.na(DISTURBANCE100m),!is.na(well_drained_soils)) %>% 
  select(LAKE_HISTORY_ID,DISTURBANCE,WaterbodyAreaSQM,WLarea,DISTURBANCE100m,well_drained_soils,septic) %>% distinct() %>% 
  distinct(LAKE_HISTORY_ID,.keep_all = TRUE)



TP_matrix<-merge(df, landcover, by = "LAKE_HISTORY_ID",all.x = TRUE)

#removing all NA values
TP_matrix<-TP_matrix %>% 
  filter(!is.na(logTP),!is.na(DISTURBANCE)) %>% distinct()

rm(list=setdiff(ls(), c("newdata",'TP_matrix')))
```



The KT Correlations of these parameters 


```{r, echo=FALSE, results='asis', message=FALSE, warning=FALSE}
library(EnvStats)

#transform for correlation analysis
matrix_TP<-TP_matrix %>% 
  select(LAKE_HISTORY_ID,depth,DISTURBANCE,DISTURBANCE100m,phosphorus,well_drained_soils,WLarea,septic) %>% distinct() %>% 
  gather(Parameter,Result,-LAKE_HISTORY_ID) %>% distinct()

#create slopes table
correlations<-data.frame(correlation=c(0,1),pair1=c("junk","junk"),pair2=c("junk","junk"))
correlations<-correlations %>% 
  mutate(pair1=as.character(pair1),pair2=as.character(pair2)) 

params<-unique(matrix_TP$Parameter)
nparams<-length(params)

for(i in 1:nparams){
  for(k in 1:nparams){
    if(params[i]!=params[k]){
    temp1<-matrix_TP %>% 
      filter(Parameter %in% c(params[i],params[k]),!is.na(Result)) %>% 
      select(LAKE_HISTORY_ID,Parameter,Result) %>% distinct() %>% 
      spread(Parameter,Result) %>% 
      filter_at(c(2,3),all_vars(!is.na(.)))

    temp3<-cor.test(temp1[,2],temp1[,3],alternative ="two.sided", method = "kendall", continuity = TRUE)
      tau<-as.data.frame(t(temp3$estimate))
      tau$pair1<-params[i]
      tau$pair2<-params[k]
      tau$pvalue<-temp3$p.value
      tau<-tau %>% 
        mutate(correlation=ifelse(pvalue>0.05,0,tau)) %>% 
        select(correlation,pair1,pair2) %>% 
        distinct()
      correlations<-merge(correlations,tau,all=TRUE)
      rm(list=setdiff(ls(), c('newdata','TP_matrix','matrix_TP','params','nparams','i','k','correlations')))
      }
  }
}

correlations2<-correlations %>% 
  filter(pair1!="junk",pair2!="junk") %>% 
  mutate(correlation=round(correlation,3)) %>% 
  spread(pair2,correlation) 

knitr::kable(correlations2, col.names = gsub("[.]", " ", names(correlations2)))



```


Conclusions:

1. Watershed disturbance has a greater correlation with phosphorus than the disturbance in the 100m buffer.

2. Septic and well drained Soils are not correlated with phosphorus so I removed them from the model


## GAM models

documentation: https://cran.r-project.org/web/packages/mgcv/mgcv.pdf

reference: https://www.mainard.co.uk/post/why-mgcv-is-awesome/#generalized-additive-models-gams

from Jacobson et al 2013

"For GAM calculations we used the mgcv package, which auto- matically selects degrees of freedom for each smooth term in the model by minimizing the generalized cross validation score. The GAM models were run using a Gaussian distri- bution and default settings.We used the GAM plot function to examine the contribution of each component’s smooth function on TP and to evaluate the statistical significance of each smooth term based on the null hypotheses that each smooth term is zero, similar to probability values of terms that would be calculated in a general linear model."

Observations:

The order of the anlyses below are necessary but confusing to understand the overall conclusions so I have summarized them here as well:


0000000. Explanations of abbreviations
			
All separated - a gam model with all parameters included
			
No 100m - a gam model with 100m buffer removed
			
No dist - a gam model with the 100m buffer and the watershed disturbance removed
			
Only_distdepth - a gam model with only disturbance and depth (as these were the most significant parameters)
			
Interact - a gam model using the interaction of the disturbance and depth


1111111.View the AIC scores first (last in the list of analyses below).

Conclusion: The model that performed best (has the lowest AIC score) was the model with only disturbance and depth.


2222222. GAM model - all parameters: 	Conclusions:
		
NOTE1: The influence of 100m disturbance and soil draining is negative which is contradictory to what we'd expect (correlation is positive). see this page which talks about this supressor effect: https://www.researchgate.net/post/What-does-it-indicating-If-there-is-positive-correlation-but-negative-regression-coefficient
		
NOTE2: 100m disturbance did not significantly contribute to the model
		
NOTE3: The concurvity of both disturbance values is over 0.8 as well. This implies that they are too similar to both include in the model. 
		
NOTE4: I ran the model without 100m disturbance and then again without watershed disturbance and the AIC was much higher when 100m disturbance was removed.
		
NOTE5: as a final test, I tried removing the 100m buffer area from the overal disturbance parameter and the results were still as notes 1-4 above
		
THEREFORE: I am removing 100m disturbance


3333333. GAM model - no watershed disturbance. 

Conclusion: The 100m disturbance is significant when we remove the influence of the watershed wide disturbance. Therefore, shoreline projects are valuable to reduce run off 

4444444. Gam Model - only watershed disturbance and depth. 	Conclusions:

The disturbance and depth model was the GAM model that performed best (lowest AIC)

When we convert the Disturbance values to their untransformed units, we can see that water quality in a lake passes the impairment threshold

5555555. Interaction

Just a really cool plot but the AIC score is high so really can't use it.

### GAM:all_parameters 


```{r}
library(mgcv)
#GAM of all parameters
all_separated <- gam(phosphorus~s(DISTURBANCE)+s(DISTURBANCE100m)+s(depth)+s(WLarea), data=TP_matrix %>% filter(!is.na(depth)))
summary(all_separated)

```

NOTE1: The influence of 100m disturbance and soil draining is negative which is contradictory to what we'd expect (correlation is positive). see this page which talks about this supressor effect: https://www.researchgate.net/post/What-does-it-indicating-If-there-is-positive-correlation-but-negative-regression-coefficient

NOTE2: 100m disturbance did not significantly contribute to the model

NOTE3: The concurvity of both disturbance values is over 0.8 as well. This implies that they are too similar to both include in the model. 

NOTE4: I ran the model without 100m disturbance and then again without watershed disturbance and the AIC was much higher when 100m disturbance was removed.

NOTE5: as a final test, I tried removing the 100m buffer area from the overal disturbance parameter and the results were still as notes 1-4 above

THEREFORE: I am removing 100m disturbance

```{r}
#GAM of only the significant parameters
#check that the k value isn't too low. none of them should be significant
gam.check(all_separated)
#check that parameters aren't too similar (worst shouldn't be >0.5)
concurvity(all_separated, full = TRUE)

#plotting
plot(all_separated,pages=1,all.terms=TRUE,residuals=TRUE,shade = TRUE)
plot(all_separated,select=1,residuals = TRUE, pch=1, cex=1,
     seWithMean = TRUE,shade = TRUE,trans = function(x){exp(coef(all_separated)[1]+x)})

```

### GAM:no100m 


```{r}
library(mgcv)
#GAM of all parameters
no100m <- gam(phosphorus~s(DISTURBANCE)+s(depth)+s(WLarea), data=TP_matrix %>% filter(!is.na(depth)))
summary(no100m)

#GAM of only the significant parameters
#check that the k value isn't too low. none of them should be significant
# gam.check(no100m)
# #check that parameters aren't too similar (worst shouldn't be >0.5)
 concurvity(no100m, full = TRUE)
# 
#plotting
plot(no100m,pages=1,all.terms=TRUE,residuals=TRUE,shade = TRUE)

plot(no100m,select=1,
     shade = TRUE,
     shift=coef(no100m)[1],
     trans=exp)
     abline(h=20, col="red")
     abline(v=.25, col="red")

```

### GAM:nodist 

removed disturbance to see if 100mdisturbance model was more significant
```{r}
library(mgcv)
#GAM of all parameters
nodist <- gam(phosphorus~s(DISTURBANCE100m)+s(depth)+s(WLarea), data=TP_matrix %>% filter(!is.na(depth)))
summary(nodist)

#GAM of only the significant parameters
#check that the k value isn't too low. none of them should be significant
# gam.check(nodist)
# #check that parameters aren't too similar (worst shouldn't be >0.5)
 concurvity(nodist, full = TRUE)
# 
#plotting
plot(nodist,pages=1,all.terms=TRUE,residuals=TRUE,shade = TRUE)

plot(nodist,select=1,
     shade = TRUE,
     shift=coef(no100m)[1],
     trans=exp)
     abline(h=20, col="red")
     abline(v=.25, col="red")

```


### GAM:only_distdepth

```{r}
library(mgcv)
#GAM of all parameters
only_distdepth <- gam(phosphorus~s(DISTURBANCE)+s(depth), data=TP_matrix %>% filter(!is.na(depth)))
summary(only_distdepth)

#GAM of only the significant parameters
#check that the k value isn't too low. none of them should be significant
gam.check(only_distdepth)
#check that parameters aren't too similar (worst shouldn't be >0.5)
concurvity(only_distdepth, full = TRUE)

#plotting
plot(only_distdepth,pages=1,all.terms=TRUE,residuals=TRUE,shade = TRUE)

plot(only_distdepth,select=1,
     shade = TRUE,
     shift=coef(only_distdepth)[1],
     trans=exp)
     abline(h=20, col="red")
     abline(v=.25, col="red")
     
plot(only_distdepth,select=3,
     shade = TRUE,residuals=TRUE,
     shift=coef(only_distdepth)[1],
     trans=exp)
     abline(h=20, col="red")

```


### Now create a smoothed interactions plot between the most significant predictors depth and disturbance

```{r}
interact <- gam(phosphorus ~ s(DISTURBANCE,depth), data = TP_matrix %>% filter(!is.na(depth)), method = "REML")
summary(interact)

#GAM of only the significant parameters
#check that the k value isn't too low. none of them should be significant
#gam.check(interact)
#check that parameters aren't too similar (worst shouldn't be >0.5)
concurvity(interact, full = TRUE)

plot(interact)
library(mgcViz)
vis.gam(interact, view = c("DISTURBANCE", "depth"),
    theta = 300, n.grid = 50, lwd = 0.4)
```

### Now compare the best fit model of those generated above

lower aic are better

```{r}

AIC(all_separated,no100m,nodist,only_distdepth,interact)


```

### Frequencies <25%


```{r}

AIC(all_separated,no100m,nodist,only_distdepth,interact)


```
